\textbf{Selective DNA and Protein Isolation from Marine Macrophyte Surfaces}

### 2020-09-14 submission/manuscript.tex
Manuscript done.

3. Department of Marine Microbiology and Biogeochemistry, Royal
Netherlands Institute for Sea Research, Utrecht University, The
Netherlands

@@

@@

Biological material (i.e.~proteins and DNA) from pelagic microbial
communities is usually isolated by collecting cells onto filters and
subsequently isolating the aimed compound (Gilbert \emph{et al.}, 2009).
If a microbial size fraction is aimed sequential filtration is applied
(Massana \emph{et al.}, 1997; Andersson \emph{et al.}, 2010). In
contrast, obtaining biological materials from microorganism inhabiting
surfaces requires either a cell detachment procedure prior to isolation
or the host material is coextracted together with the targeted material.
Methods for separating microbial cells form the host include shaking of
host tissue (Gross \emph{et al.}, 2003; Nõges \emph{et al.}, 2010),
scraping of macrophyte surfaces (Uku \emph{et al.}, 2007) or the
application of ultrasonication (Weidner \emph{et al.}, 1996; Cai
\emph{et al.}, 2014). It was shown that shaking alone is not sufficient
to remove microbial cells, at least from plant root surfaces
(Richter-Heitmann \emph{et al.}, 2016). Manual separation methods, such
as scraping and brushing, are time consuming and subjective, as the
detachment efficiency depends on host tissue and the person performing
the procedure (Cai \emph{et al.}, 2014). Ultrasonication was proposed as
an alternative method as it is providing better results in terms of
detachment efficiency (Cai \emph{et al.}, 2014; Richter-Heitmann
\emph{et al.}, 2016). The downside of this procedure is that complete
cell removal was still not obtained and tissue disruption was observed
especially after the application of probe ultrasonication
(Richter-Heitmann \emph{et al.}, 2016). An alternative to these cell
detachment procedures is the isolation of targeted epiphytic compounds
together with host materials (Staufenberger \emph{et al.}, 2008; Jiang
\emph{et al.}, 2015). This procedure can lead to problems in the
following processing steps such as mitochondrial and chloroplast 16S
rRNA sequence contaminations from the host (Longford \emph{et al.},
2007; Staufenberger \emph{et al.}, 2008). In addition, when performing
metagenomics and metaproteomics host material can cause biased results
towards more abundant host DNA and proteins.

An alternative to these procedures is a direct isolation of the targeted
material by incubating macrophyte tissues in an extraction buffer. After
the incubation is done the undisrupted host tissue is removed and the
isolation procedure continues omitting host material contaminations. To
our knowledge the only procedure describing a direct and selective
epiphytic DNA isolation from the surfaces of marine macrophytes was
provided by Burke \emph{et al.} (2009). In contrast to previously
described methods this protocol enables an almost complete removal of
the surface community and was used for 16S rRNA gene clone libraries
construction (Burke and Thomas \emph{et al.}, 2011) and metagenomes
sequencing (Burke and Peter Steinberg \emph{et al.}, 2011). This method,
thought providing a selective isolation procedure, is using in the
extraction buffer a rapid multienzyme cleaner (3M) that is not worldwide
available and whose composition is not know (Burke \emph{et al.}, 2009).
Also to our knowledge, no selective isolation protocol for proteins from
epiphytic communities inhabiting marine macrophytes was established.

In the present study, we adapted a protocol widely used for DNA
isolation from filters (Massana \emph{et al.}, 1997) and a protocol used
for protein isolation from sediments (Chourey \emph{et al.}, 2010;
Hultman \emph{et al.}, 2015) for selective extractions of DNA and
proteins from epiphytic communities inhabiting the surfaces of two
marine macrophytes: the seagrass \emph{Cymodocea nodosa} and the alga
\emph{Caulerpa cylindracea}. In addition, we tested the removal
efficiency of the protocol and the suitability of obtained DNA and
proteins for 16S rRNA sequencing and metaproteomics.

@@

To assess the removal efficiency of DNA and protein isolation procedures
leaves and thalli were stained with SYBR Green I and examined under a
confocal microscope before and after treatments were performed. As it is
known that marine macrophytes are harboring more algal epiphytes during
Autumn and Winter months (Reyes and Sansón, 2001), the sampling was
performed in two contrasting seasons, in December 2017 when higher
abundances of algal epiphytes are expected, and in June 2018, when lower
abundances could be expected. In addition, to further test the isolation
protocols on different macrophyte species, DNA and proteins were
isolated from the surfaces of two macrophytes, the seagrass \emph{C.
nodosa} and the macroalga \emph{C. cylindracea}. \emph{C. nodosa} was
sampled in a monospecific meadow (Bay of Saline) and \emph{C.
cylindracea} invaded meadow (Bay of Funtana), while samples of \emph{C.
cylindracea} originated from a monospecific settlement and a \emph{C.
cylindracea} invaded \emph{C. nodosa} meadow in the Bay of Funtana.
\emph{C. nodosa} June samples from the Bay of Saline were missing as the
meadow went to a decline from Spring 2018 (M. Najdek, personal
communication). Procedures developed for DNA and protein isolation
showed an almost complete removal of the surface community on both,
samples collected in December 2017 (\autoref{confocal_december}) and
June 2018 (\autoref{confocal_june}). In addition, a similar removal
efficiency was observed for communities associated with \emph{C. nodosa}
and \emph{C. cylindracea}. Also, no effect of station, settlement or
isolation procedure (DNA or protein) on the removal efficiency was
observed (Figs. \ref{confocal_december} and \ref{confocal_june}).

To evaluate if the obtained DNA is suitable to determine the microbial
community structure an Illumina sequencing of the V4 16S rRNA region was
performed. Sequencing yielded a total of 336,937 sequences after quality
curation and exclusion of eukaryotic, mitochondrial and no relative
sequences. The number of sequences classified as chloroplast was 97,328
(28.9 \si{\percent})˛ After the exclusion of these sequences the total
number of retrieved reads dropped to 239,609, ranging from 22,587 and
52,958 sequences per sample (\autoref{supp-nseq_notus}). Even when the
highest sequencing effort was applied the rarefaction curves did not
level off that is commonly observed in high-throughput 16S rRNA amplicon
sequencing procedures (\autoref{supp-rarefaction}). Sequence clustering
at a similarity level of 97 \si{\percent} yielded a total of 8,360
different OTUs. Taxonomic classification of reads allowed the
determination of the macrophyte associated epiphytic community that was
mainly composed of: \emph{Alphaproteobacteria} (14.9 ± 3.5
\si{\percent}), \emph{Bacteroidota} (12.5 ± 2.4 \si{\percent}),
\emph{Gammaproteobacteria} (11.6 ± 4.3 \si{\percent}),
\emph{Desulfobacterota} (7.8 ± 7.4 \si{\percent}), \emph{Cyanobacteria}
(6.5 ± 4.7 \si{\percent}) and \emph{Planctomycetota} (2.9 ± 1.7
\si{\percent}) (\autoref{community}).

Primers used in this study, as in many other 16S rRNA amplicon
procedures, also amplify chloroplast 16S rRNA genes. We observed a high
proportion of chloroplast sequences in all analyzed samples (33.4 ± 9.4
\si{\percent}) (\autoref{community}). To determine if chloroplast
sequences originate from hosts or eukaryotic epiphytic organisms, we
exported SILVA classified chloroplast sequences and reclassified them
using the RDP (Ribosomal Database Project) training set that allows for
a more detailed chloroplast classification. The largest proportion of
sequences was classified as Bacillariophyta (89.7 ± 5.7 \si{\percent})
indicating that the DNA removal procedure did not coextract larger
quantities of host DNA (\autoref{chloroplast}).

@@

@@

Leaves of \emph{C. nodosa} were sampled in a \emph{C. nodosa} meadow in
the Bay of Saline (\ang{45;7;5} N, \ang{13;37;20} E) and in a \emph{C.
nodosa} meadow invaded by \emph{C. cylindracea} in the proximity of the
village of Funtana (\ang{45;10;39} N, \ang{13;35;42} E). Thalli of
\emph{C. cylindracea} were sampled in the same \emph{C. nodosa} invaded
meadow in Funtana and on a locality of only \emph{C. cylindracea}
located in the proximity of the invaded meadow. Leaves and thalli were
collected on the same day in two contrasting seasons, on 4 December 2017
and 18 June 2018. During spring 2018 the \emph{C. nodosa} meadow in the
Bay of Saline decayed to an extent that no leaves could be retrieved
(Najdek \emph{et al.}, unpublished data). Leaves and thalli were
collected by diving and transported to the laboratory in containers
placed on ice and filled with site seawater. Upon arrival to the
laboratory, \emph{C. nodosa} leaves were cut into sections of 1 -- 2
\si{\cm}, while \emph{C. cylindracea} thalli were cut into 5 -- 8
\si{\cm} long sections. Leaves and thalli were washed three times with
sterile artificial seawater (ASW) to remove loosely attached microbial
cells.

@@

The DNA was isolated according to the protocol for isolation from
filters described in Massana \emph{et al.} (1997). This protocol was
modified and adapted for DNA isolation from microbial communities from
macrophyte surfaces as described below. 5 \si{\ml} of lysis buffer (40
\si{\milli\Molar} EDTA, 50 \si{\milli\Molar} Tris-HCl, 0.75 \si{\Molar}
sucrose; pH 8.3) was added to 1 \si{\g} wet weight of leaves or \si{\g}
wet-weight of thalli. Lysozyme was added (final concentration 1
\si{\mg\per\ml}) and the mixture was incubated at 37 \si{\degreeCelsius}
for 30 \si{\minute}. Subsequently, proteinase K (final concentration 0.5
\si{\mg\per\ml}) and SDS (final concentration 1 \si{\percent}) were
added and the samples were incubated at 55 \si{\degreeCelsius} for 2
\si{\hour}. Following the incubation, tubes were vortexed for 10
\si{\minute} and the mixture containing lyzed epiphytic cells was
separated from host leaves or thalli by transferring the solution into a
clean tube. The lysate was extracted twice with a mixture of
phenol:chloroform:isoamyl alcohol (25:24:1; pH 8) and once with
chloroform:isoamyl alcohol (24:1). After each organic solvent mixture
addition tubes were slightly vortexed and centrifuged at 4,500 × g for
10 \si{\minute}. Following each centrifugation aqueous phases were
retrieved. After the final extraction 1/10 of chilled 3 \si{\Molar}
sodium acetate (pH 5.2) was added. DNA was precipitated by adding 1
volume of chilled isopropanol, incubating the mixtures overnight at
\num{-20} °C and centrifuging at 16,000 × g and 4 \si{\degreeCelsius}
for 20 \si{\minute}. The pellet was washed twice with 1 \si{\ml} of
chilled 70 \si{\percent} ethanol and centrifuged after each washing step
at 20,000 × g and 4 \si{\degreeCelsius} for 10 \si{\minute}. After the
first washing step duplicate pellets form the same sample were pooled
and transferred to a clean 1.5 \si{\ml} tube. The dried pellet was
resuspended in 100 \si{\ul} of deionized water.

@@

Four samples were selected and sent on dry ice to IMGM Laboratories,
Martinsried, Germany for metagenomic sequencing. DNA was purified using
AMPure XP Beads (Beckman Coulter, USA) applying a bead:DNA ratio of 1:1
(v/v), quantified with a Qubit dsDNA HS Assay Kit (Thermo Fisher
Scientific, USA) and check for integrity on a 1 \si{\percent} agarose
gel. Metagenomic sequencing libraries were generated from 300 \si{\ng}
of input DNA using a NEBNext Ultra II FS DNA Library Prep Kit for
Illumina (New England Biolabs, USA) according to the manufacturer's
instructions. Fragments were selected (500 -- 700 bp) using AMPure XP
Beads, PCR enriched for 3 -- 5 cycles and quality controlled. Libraries
generated from different DNA samples were pooled and sequenced on an
Illumina NovaSeq 6000 sequencing system (2 × 250 bp).

>>>>

3. NIOZ, Department of Marine Microbiology and Biogeochemistry, Royal
Netherlands Institute for Sea Research, Utrecht University, The
Netherlands

@@

Studies of unculturable microbes often combine methods based on DNA,
such as 16S rRNA sequencing and metagenomics, with methods that allow an
insight into the metabolic status, such as metaproteomics. To apply
these procedures to the microbial community inhabiting the surfaces of
marine macrophytes it is advisable to perform, prior to the analysis, a
selective DNA and protein isolation so that the host material, present
in higher quantities, is not hampering the analysis. Two protocols, for
DNA and protein isolation, were adapted for selective extractions of DNA
and proteins from epiphytic communities inhabiting the surfaces of two
marine macrophytes: the seagrass \emph{Cymodocea nodosa} and the
macroalga \emph{Caulerpa cylindracea}. Protocols showed an almost
complete removal of the epiphytic community regardless of the sampling
season, station, settlement or host species. The obtained DNA was
suitable for metagenomic and 16S rRNA sequencing, while isolated
proteins could be identified by mass spectrometry showing that protocols
can be used in 16S rRNA, metagenomic and metaproteomic analysis. Low
presence of host DNA and proteins, observed in isolated samples,
indicated a selective nature of protocols. Furthermore, the procedures
are based on universally available laboratory chemicals making the
protocols widely applicable. Taken together, the adapted protocols
ensure an almost complete removal of the macrophyte epiphytic community,
are selective for microbes inhabiting macrophyte surfaces and are
providing DNA and proteins applicable in 16S rRNA sequencing,
metagenomics and metaproteomics.

@@

Biological material (i.e.~proteins and DNA) from pelagic microbial
communities is usually isolated by collecting cells onto filters and
subsequently isolating the aimed compound (Gilbert \emph{et al.}, 2009).
If a specific microbial size fraction is aimed sequential filtration is
applied (Massana \emph{et al.}, 1997; Andersson \emph{et al.}, 2010). In
contrast, obtaining biological materials from microorganism inhabiting
surfaces requires either a cell detachment procedure prior to isolation
or the host material is coextracted together with the targeted material.
Methods for separating microbial cells form the host include shaking of
host tissue (Gross \emph{et al.}, 2003; Nõges \emph{et al.}, 2010),
scraping of macrophyte surfaces (Uku \emph{et al.}, 2007) or the
application of ultrasonication (Weidner \emph{et al.}, 1996; Cai
\emph{et al.}, 2014). It was shown that shaking alone is not sufficient
to remove microbial cells, at least from plant root surfaces
(Richter-Heitmann \emph{et al.}, 2016). Manual separation methods, such
as scraping and brushing, are time consuming and subjective, as the
detachment efficiency depends on host tissue and the person performing
the procedure (Cai \emph{et al.}, 2014). Ultrasonication was proposed as
an alternative method as it is providing better results in terms of
detachment efficiency (Cai \emph{et al.}, 2014; Richter-Heitmann
\emph{et al.}, 2016). The downside of this procedure is that complete
cell removal was still not obtained and tissue disruption was observed
especially after the application of probe ultrasonication
(Richter-Heitmann \emph{et al.}, 2016). An alternative to these cell
detachment procedures is the isolation of targeted epiphytic compounds
together with host materials (Staufenberger \emph{et al.}, 2008; Jiang
\emph{et al.}, 2015). This procedure can lead to problems in the
following processing steps such as mitochondrial and chloroplast 16S
rRNA sequence contaminations from the host (Longford \emph{et al.},
2007; Staufenberger \emph{et al.}, 2008). In addition, when performing
metagenomics and metaproteomics host material can cause biased results
towards more abundant host DNA and proteins.

An alternative to these procedures is a direct isolation of the targeted
material by incubating macrophyte tissues in an extraction buffer. After
the incubation, the undisrupted host tissue is removed and the isolation
procedure continues omitting host material contaminations. To our
knowledge the only procedure describing a direct and selective epiphytic
DNA isolation from the surfaces of marine macrophytes was provided by
Burke \emph{et al.} (2009). In contrast to previously described methods
this protocol enables an almost complete removal of the surface
community and it was used for 16S rRNA gene clone library construction
(Burke and Thomas \emph{et al.}, 2011) and metagenome sequencing (Burke
and Peter Steinberg \emph{et al.}, 2011). This method, thought providing
a selective isolation procedure, uses a rapid multienzyme cleaner (3M)
that is not available worldwide and without a known composition (Burke
\emph{et al.}, 2009). Also to our knowledge, no selective isolation
protocol for proteins from epiphytic communities inhabiting marine
macrophytes was established.

In the present study, we adapted a protocol widely used for DNA
isolation from filters (Massana \emph{et al.}, 1997) and a protocol used
for protein isolation from sediments (Chourey \emph{et al.}, 2010;
Hultman \emph{et al.}, 2015) for selective extractions of DNA and
proteins from epiphytic communities inhabiting the surfaces of two
marine macrophytes: the seagrass \emph{Cymodocea nodosa} and the
macroalga \emph{Caulerpa cylindracea}. In addition, we tested the
removal efficiency of the protocol and the suitability of obtained DNA
and proteins for 16S rRNA sequencing, metagenomics and metaproteomics.

@@

To assess the removal efficiency of DNA and protein isolation procedures
leaves and thalli were stained with SYBR Green I and examined under a
confocal microscope before and after treatments were performed. To test
the isolation protocols on different macrophyte species, DNA and
proteins were isolated from the surfaces of two macrophytes, the
seagrass \emph{C. nodosa} and the macroalga \emph{C. cylindracea}.
\emph{C. nodosa} was sampled in a monospecific meadow (Bay of Saline)
and \emph{C. cylindracea} invaded meadow (Bay of Funtana), while samples
of \emph{C. cylindracea} originated from a monospecific settlement and a
\emph{C. cylindracea} invaded \emph{C. nodosa} meadow in the Bay of
Funtana. In addition, the sampling was performed in two contrasting
seasons, in December 2017 and in June 2018, to test the impact of
seasonal conditions on the removal efficiency. Procedures developed for
DNA and protein isolation showed an almost complete removal of the
surface community of both, \emph{C. nodosa} and \emph{C. cylindracea}.
In addition, a similar removal efficiency was observed for communities
sampled in December 2017 (\autoref{confocal_december}) and June 2018
(\autoref{confocal_june}). Also, no effect of station, settlement or
isolation procedure (DNA or protein) on the removal efficiency was
observed (Figs. \ref{confocal_december} and \ref{confocal_june}).

To evaluate if the obtained DNA is suitable to determine the microbial
community structure an Illumina sequencing of the V4 16S rRNA region was
performed. Sequencing yielded a total of 336,937 sequences after quality
curation and exclusion of eukaryotic, mitochondrial and no relative
sequences. The number of sequences classified as chloroplast was 97,328.
After excluding these sequences the total number of retrieved reads
dropped to 239,609, ranging from 22,587 to 52,958 sequences per sample
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off which is commonly
observed in high-throughput 16S rRNA amplicon sequencing procedures
(\autoref{supp-rarefaction}). Sequence clustering at a similarity level
of 97 \si{\percent} yielded a total of 8,360 different OTUs. Taxonomic
classification of reads allowed for a macrophyte associated epiphytic
community determination that was mainly composed of:
\emph{Alphaproteobacteria} (14.9 ± 3.5 \si{\percent}),
\emph{Bacteroidota} (12.5 ± 2.4 \si{\percent}),
\emph{Gammaproteobacteria} (11.6 ± 4.3 \si{\percent}),
\emph{Desulfobacterota} (7.8 ± 7.4 \si{\percent}), \emph{Cyanobacteria}
(6.5 ± 4.7 \si{\percent}) and \emph{Planctomycetota} (2.9 ± 1.7
\si{\percent}) (\autoref{community}).

Primers used in this study, as in many other 16S rRNA amplicon
procedures, also amplified chloroplast 16S rRNA genes. We observed a
high proportion of chloroplast sequences in all analyzed samples (33.4 ±
9.4 \si{\percent}) (\autoref{community}). To determine if chloroplast
sequences originate from hosts or eukaryotic epiphytic organisms, we
exported SILVA classified chloroplast sequences and reclassified them
using the RDP (Ribosomal Database Project) training set that allows for
a more detailed chloroplast classification. The largest proportion of
sequences were classified as Bacillariophyta (89.7 ± 5.7 \si{\percent})
indicating that the DNA removal procedure did not coextract larger
quantities of host DNA (\autoref{chloroplast}). Chloroplast sequences
classified as Streptophyta constituted 3.3 ± 2.8 \si{\percent} of all
chloroplast sequences originating from \emph{C. nodosa} samples, while
sequences classified as Chlorophyta comprised only 0.02 ± 0.01
\si{\percent} of all chloroplast sequences associated with \emph{C.
cylindracea} samples.

To determine if the extracted DNA can be used for metagenomic sequencing
four samples containing epiphytic DNA were selected and shotgun
sequenced using an Illumina platform. Metagenomic sequencing yielded
from 207,149,524 to 624,029,930 sequence pairs
(\autoref{supp-metagenomic_statistics}). Obtained sequences were
successfully assembled into contigs whose L50 ranged from 654 to 1,011
bp. In addition, predicted coding sequences were successfully
functionally annotated (9,066,667 -- 20,256,215 annotated sequences;
\autoref{cog}) and taxonomically classified. Functional annotation
allowed for an assessment of the relative contribution of each COG
(Clusters of Orthologous Groups) functional category to the total number
of annotated coding sequences (\autoref{cog}a). Functional categories
containing the highest number of sequences were: C (Energy production
and conversion), E (Amino acid transport and metabolism), M (Cell
wall/membrane/envelope biogenesis), L (Replication, recombination and
repair) and P (Inorganic ion transport and metabolism). If host DNA was
coextracted with epiphytic it should be detected in larger proportions
in sequenced metagenomes. Indeed, no higher proportions of coding
sequences classified into phylum Streptophyta or Chlorophyta were
detected (\autoref{supp-metagenomic_taxonomy}). Sequenced metagenomic
DNA originating from the surface of \emph{C. nodosa} contained 1.3
\si{\percent} of coding sequences classified into the phylum
Streptophyta in December 2017 and 0.7 \si{\percent} in June 2018.
Furthermore, the summed RPKM (Reads Per Kilobase Million) of these
sequences constituted 1.7 \si{\percent} of total RPKM of all
successfully classified sequences in December 2017 and 1.1 \si{\percent}
in June 2018. Similar low proportions of host coding sequences were
detected in metagenomic samples originating from the surfaces of
\emph{C. cylindracea}. Of all successfully classified coding sequences
0.2 \si{\percent} were classified into Chlorophyta in December 2017 and
0.1 \si{\percent} in June 2018. A relatively higher proportion of these
sequences' RPKM than in the case of \emph{C. nodosa} was observed,
indicating a higher coextraction of host DNA in \emph{C. cylindracea}.
In December, the proportion of RPKM of sequences classified into
Chlorophyta was 8.2 \si{\percent}, while in June 2018 it reached 13.6
\si{\percent}.

To evaluate if the procedure for protein extraction is suitable for
metaproteomic analysis, obtained proteins were trypsin digested and
sequenced using a mass spectrometer. Obtained MS/MS spectra were
searched against a protein database from sequenced metagenomes. From
14,219 to 16,449 proteins were identified in isolated protein samples
(\autoref{cog}b). In addition, successful identification of proteins
allowed for an assessment of the relative contribution of each COG
functional category to the total number of identified proteins
(\autoref{cog}b). Functional categories containing the highest number of
identified proteins were: C (Energy production and conversion), G
(Carbohydrate transport and metabolism), P (Inorganic ion transport and
metabolism), O (Posttranslational modification, protein turnover,
chaperones) and E (Amino acid transport and metabolism). Isolated
proteins could derive from epiphytic organisms inhabiting the macrophyte
surface or from macrophyte tissue underlying them. The contribution of
proteins originating from host tissues was evaluated by identifying all
the proteins predicted to belong to any taxonomic group within the phyla
Streptophyta and Chlorophyta and by calculating their contribution to
the number and abundance of all identified proteins. On average,
proteins isolated from the surface of \emph{C. nodosa} contained 1.8 ±
0.06 \si{\percent} of proteins associated with Streptophyta,
contributing to 2.2 ± 0.8 \si{\percent} of total proteins. Similar to
metagenomes, proteins associated with Chlorophyta contributed more to
\emph{C. cylindracea} than proteins associated with Streptophyta to
\emph{C. nodosa}. Chlorophyta associated proteins comprised 5.2 ± 0.06
\si{\percent} of all identified proteins in \emph{C. cylindracea},
contributing to 19.2 ± 1.5 \si{\percent} of all protein abundances.

@@

The study of marine macrophyte epiphytic communities using culture
independent techniques, such as 16S rRNA analysis, metagenomics and
metaproteomics, requires an initial step of biological material
isolation. Methods that have been developed for selective isolation of
epiphytic biological material could be divided into two groups: (i)
procedures involving, prior to extraction, a cell detachment step such
as shaking (Gross \emph{et al.}, 2003; Nõges \emph{et al.}, 2010),
scraping (Uku \emph{et al.}, 2007) and ultrasonication (Weidner \emph{et
al.}, 1996; Cai \emph{et al.}, 2014) and (ii) procedures involving a
host tissue incubation aiming at direct lysis of epiphytic microbial
cells (Burke \emph{et al.}, 2009). Protocols that include a cell
detachment procedure do not provide complete cell removal (Cai \emph{et
al.}, 2014; Richter-Heitmann \emph{et al.}, 2016) and in the case of
probe ultrasonication tissue disruption can also be observed
(Richter-Heitmann \emph{et al.}, 2016). To avoid these problems a
procedure for DNA isolation involving host tissue incubation in the
extraction buffer and direct epiphytic cell lysis on the macrophyte
surface was developed (Burke \emph{et al.}, 2009). Though providing an
almost complete cell removal and selective DNA isolation this protocol
uses in the extraction buffer a rapid multienzyme cleaner (3M) not
available worldwide and without a known composition (Burke \emph{et
al.}, 2009). To circumvent these problems we developed and tested two
protocols for selective DNA and protein isolation from marine macrophyte
epiphytic communities.

To test if the developed DNA and protein isolation procedures could be
applied on a variety of macrophyte species we selected \emph{C. nodosa}
and \emph{C. cylindracea} as representatives of seagrass and macroalgal
species, on which the procedures were tested. These species especially
differ morphologically. While \emph{C. nodosa} leaves are flat, \emph{C.
cylindracea} thallus is composed of uneven surfaces (Kuo and den Hartog,
2001; Verlaque \emph{et al.}, 2003). The developed procedures showed an
almost complete removal of epiphytic cells from the surfaces of both
species comparable to the result of Burke \emph{et al.} (2009) and
indicating that structural differences do not impact the removal
efficiency. In addition, isolation protocols were tested in two
contrasting season as it is known that macrophytes are harboring more
algal epiphytes during autumn and winter (Reyes and Sansón, 2001). No
differences in the removal efficiency was observed between seasons
suggesting that protocols can be used on macrophyte samples retrieved
throughout the year. Also, no removal differences were observed on
samples derived from the same host but from different localities or
settlements demonstrating the stability of the protocol in cell removal
efficiency.

Successful amplification and sequencing of the V4 16S rRNA gene region
proved that the isolated DNA can be used to estimate the microbial
epiphytic diversity. Taxonomic groups detected in this step can also be
often found in epiphytic communities associated with other macrophytes
(Burke and Thomas \emph{et al.}, 2011; Morrissey \emph{et al.}, 2019). A
problem often encountered in studies focusing on epiphytic communities
is the presence of large proportions of chloroplast 16S rRNA sequences
in the pool of amplified molecules, especially if the epiphytic DNA was
isolated without prior selection (Staufenberger \emph{et al.}, 2008).
These sequences can derive from host chloroplasts or from eukaryotic
epiphytic chloroplast DNA. Although the proportion of obtained
chloroplast 16S rRNA sequences in our samples was substantial they
derived almost exclusively from eukaryotic epiphytes. High proportion of
chloroplast 16S rRNA sequences in studies applying selective procedures
that include direct cellular lysis on host surfaces were observed before
(Michelou \emph{et al.}, 2013). It is possible that chloroplast specific
sequences even in these studies are originating from eukaryotic
epiphytic cells and not host chloroplast. Indeed, it is common during
16S rRNA profiling of pelagic microbial communities to observe high
proportions of chloroplast sequences (Gilbert \emph{et al.}, 2009;
Korlević \emph{et al.}, 2016). In addition, a very low proportion of
chloroplast 16S rRNA sequences in samples originating from \emph{C.
cylindracea} in comparison to \emph{C. nodosa} could be explained by the
presence of three introns in the gene for 16S rRNA in some members of
the genus \emph{Caulerpa} that could hamper the amplification process
(Lam and Lopez-Bautista, 2016).

Beside 16S rRNA sequencing high quality DNA is needed for metagenomics.
The obtained number of metagenomic sequences and assembly statistics
were comparable to metagenomes and metatranscriptomes derived from
similar surface associated communities (Crump \emph{et al.}, 2018; Cúcio
\emph{et al.}, 2018). In addition, functional annotation of predicted
coding sequences to COG functional categories showed that the obtained
metagenomes can be used to determine the metabolic capacity of surface
associated communities (Leary \emph{et al.}, 2014; Cúcio \emph{et al.},
2018). The proportion of coding sequences, including their RPKM,
originating from \emph{C. nodosa} metagenomes and classified as
Streptophyta was low indicating a good selectivity of the isolation
procedure towards epiphytic cells. In the case of DNA samples isolated
from the surface of \emph{C. cylindracea} the proportion of Chlorophyta
coding sequences was also low but their RPKM was higher than in the case
of \emph{C. nodosa}. One of the causes for this elevated RPKM of
Chlorophyta sequences in \emph{C. cylindracea} could lay in tissue
structure differences between these two host species. While \emph{C.
nodosa} leaves are composed of individual cells, the thallus of \emph{C.
cylindracea} is, like in other siphonous algal species, composed of a
single large multinucleate cell (Coneva and Chitwood, 2015). The absence
of individual cells in \emph{C. cylindracea} could cause a leakage of
genetic material into the extraction buffer causing an elevated presence
of host sequences in metagenomic data.

Metagenomic analysis reveals the potential metabolic capacity of
microbial communities, while to obtain an insight into the metabolic
status of uncultivated members, a metaproteomic approach is required
(Saito \emph{et al.}, 2019). The applied protocol for epipyhtic protein
isolation followed by a metaproteomic analysis identified between 14,219
and 16,449 proteins that is higher than previously reported for
e.g.~soils (Chourey \emph{et al.}, 2010; Hultman \emph{et al.}, 2015),
seawater (Williams \emph{et al.}, 2012) and biofilms (Leary \emph{et
al.}, 2014). The functional annotation of identified proteins into COG
functional categories showed that the protein isolation procedure can be
used to assess the metabolic status of the macrophyte epiphytic
community (Leary \emph{et al.}, 2014). Similar to metagenomes, the
number of identified proteins, including their abundances, associated
with Streptophyta in \emph{C. nodosa} samples were low indicating that
the procedure is selective for epiphytic cell proteins. In addition, a
higher number of identified proteins, and especially their abundances,
associated with Chlorophyta was observed in \emph{C. cylindracea}
samples. The cause of this elevated presence of Chlorophyta associated
proteins can be, as in the case of the DNA isolation procedure,
explained by the absence of individual cells in this siphonous alga
(Coneva and Chitwood, 2015).

In conclusion, the developed protocols for DNA and protein isolation
from macrophyte surfaces are providing an almost complete removal of the
epiphytic community and are shown to ensure removal from both, \emph{C.
nodosa} and \emph{C. cylindracea}, in different seasons. Also, the
obtained DNA and proteins are suitable for 16S rRNA sequencing,
metagenomics and metaproteomics, while the obtained material contains
low quantities of host DNA or proteins making the procedures epiphyte
selective. Furthermore, the procedures are based on universally
available laboratory chemicals making the protocols widely applicable.

@@

Leaves of \emph{C. nodosa} were sampled in a \emph{C. nodosa} meadow in
the Bay of Saline (\ang{45;7;5} N, \ang{13;37;20} E) and in a \emph{C.
nodosa} meadow invaded by \emph{C. cylindracea} in the proximity of the
village of Funtana (Bay of Funtana; \ang{45;10;39} N, \ang{13;35;42} E).
Thalli of \emph{C. cylindracea} were sampled in the same \emph{C.
nodosa} invaded meadow in the Bay Funtana and on a locality of only
\emph{C. cylindracea} located in the proximity of the invaded meadow.
Leaves and thalli for 16S rRNA analysis, metagenomics and metaproteomics
were collected in two contrasting seasons, on 4 December 2017 (16S rRNA
analysis and metaproteomics), 14 December 2017 (metagenomics) and 18
June 2018 (16S rRNA analysis, metagenomics and metaproteomics). During
spring 2018 the \emph{C. nodosa} meadow in the Bay of Saline decayed to
an extent that no leaves could be retrieved (Najdek \emph{et al.},
2020). In addition, as not enough DNA for both metagenomic and 16S RNA
analysis were obtained during the sampling on 4 December 2017, an
additional sampling on 14 December 2017 was carried out in the Bay of
Funtana. Leaves and thalli were collected by diving and transported to
the laboratory in containers placed on ice and filled with site
seawater. Upon arrival to the laboratory, \emph{C. nodosa} leaves were
cut into sections of 1 -- 2 \si{\cm}, while \emph{C. cylindracea} thalli
were cut into 5 -- 8 \si{\cm} long sections. Leaves and thalli were
washed three times with sterile artificial seawater (ASW) to remove
loosely attached microbial cells.

@@

The DNA was isolated according to the protocol for isolation from
filters described in Massana \emph{et al.} (1997). This protocol was
modified and adapted for microbial DNA isolation from macrophyte
surfaces as described below. 5 \si{\ml} of lysis buffer (40
\si{\milli\Molar} EDTA, 50 \si{\milli\Molar} Tris-HCl, 0.75 \si{\Molar}
sucrose; pH 8.3) was added to 1 \si{\g} wet weight of leaves or 2
\si{\g} wet-weight of thalli. For every sample duplicate isolations were
performed. Lysozyme was added (final concentration 1 \si{\mg\per\ml})
and the mixture was incubated at 37 \si{\degreeCelsius} for 30
\si{\minute}. Subsequently, proteinase K (final concentration 0.5
\si{\mg\per\ml}) and SDS (final concentration 1 \si{\percent}) were
added and the samples were incubated at 55 \si{\degreeCelsius} for 2
\si{\hour}. Following the incubation, tubes were vortexed for 10
\si{\minute} and the mixture containing lyzed epiphytic cells was
separated from host leaves or thalli by transferring the solution into a
clean tube. The lysate was extracted twice with a mixture of
phenol:chloroform:isoamyl alcohol (25:24:1; pH 8) and once with
chloroform:isoamyl alcohol (24:1). After each organic solvent mixture
addition tubes were slightly vortexed and centrifuged at 4,500 × g for
10 \si{\minute}. Following each centrifugation aqueous phases were
retrieved. After the final extraction 1/10 of chilled 3 \si{\Molar}
sodium acetate (pH 5.2) was added. DNA was precipitated by adding 1
volume of chilled isopropanol, incubating the mixtures overnight at
\num{-20} °C and centrifuging at 16,000 × g and 4 \si{\degreeCelsius}
for 20 \si{\minute}. The pellet was washed twice with 1 \si{\ml} of
chilled 70 \si{\percent} ethanol and centrifuged after each washing step
at 20,000 × g and 4 \si{\degreeCelsius} for 10 \si{\minute}. After the
first washing step duplicate pellets form the same sample were pooled
and transferred to a clean 1.5 \si{\ml} tube. The dried pellet was
resuspended in 100 \si{\ul} of deionized water.

@@

Four DNA samples were selected and sent on dry ice to IMGM Laboratories,
Martinsried, Germany for metagenomic sequencing. DNA was purified using
AMPure XP Beads (Beckman Coulter, USA) applying a bead:DNA ratio of 1:1
(v/v), quantified with a Qubit dsDNA HS Assay Kit (Thermo Fisher
Scientific, USA) and check for integrity on a 1 \si{\percent} agarose
gel. Metagenomic sequencing libraries were generated from 300 \si{\ng}
of input DNA using a NEBNext Ultra II FS DNA Library Prep Kit for
Illumina (New England Biolabs, USA) according to the manufacturer's
instructions. Fragments were selected (500 -- 700 bp) using AMPure XP
Beads, PCR enriched for 3 -- 5 cycles and quality controlled. Libraries
generated from different DNA samples were pooled and sequenced on an
Illumina NovaSeq 6000 sequencing system (2 × 250 bp).
